---
title: "PTMC_usage"
author: "David Hodgson"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PTMC_usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup}
library(tidyverse)
library(ptmc)
library(coda)
```

# Parallel Tempering Monte Carlo

## Model 

The example model is taken from Ben Lambert's fantastic book, "A Student's Guide to Bayesian Statistics," [Questions 13.3](https://benlambertdotcom.files.wordpress.com/2018/08/bayesianbook_problemsanswers_final.pdf#page=124). Breifly, I assume the random count variable $X_t$, the number of mosquitos caught during day $t$, is Poission, such that

\begin{equation}
\begin{split}
X_t &\sim \textit{Poisson}(\lambda)\\ 
\lambda &= 1000 \times \exp(-\mu t)\psi
\end{split}
\end{equation}

where $\mu$ is a constant mortality hazard rate and $\psi$ is the daily recapture probability with priors distributions of $\mu \sim \Gamma(2,20)$ and $\psi \sim \textit{Beta}(2, 40)$. The data $x_t$, the number of mosquitos caught on day $t$ is given in the RWM_mosquito.csv file but can be obtained [here](https://benlambertdotcom.files.wordpress.com/2018/08/all_data.zip). 


## Implementation  

### Create the model
The `model` variable should be a list with three functions, `gen_init`, `eval_lpr`, and `eval_ll` and a vector string with the parameter names used in the calibration, `par_names`.

#### vector string `parnames` 
A vector string of the parameter names 

#### func `gen_init` 
A function which generates the original samples (usually the prior distributions)

* (no arguments)
* @return vector of values 

#### func `eval_ll` 
A function which generates the log likelihood for a set of parameter values

* @param data, data needed to calcualte the log likelihood
* @param params, parameter values
* @return log likelihood 

#### func `eval_lpr`
A fucntion which calcualte the log prior for a set of parameter values

* @param params, parameter values
* @return log prior 


```{r model definition}
# model is a list of three functions
model <- list(

  par_names = c("mu","psi"),

  # Generate the initial step of Markov chain
  gen_init = function() {
    c(rgamma(1, 2, 20), rbeta(1, 2, 40))
  },

  # Evaluate the log prior
  eval_lpr = function(params) {
    if (params[1] < 0 || params[2] < 0 || params[1] > 1 || params[2] > 1)
      return(-Inf)

    lpr <- 0
    lpr <- lpr + log(pgamma( params[1], 2,20))
    lpr <- lpr + log(pbeta( params[2], 2,40))
    lpr
  },

  # Evaluate the log likelihood
  eval_ll = function(data, params) {
    y <- data$y
    mu <- params[1]
    psi <- params[2]
    ll <- 0;
    for (t in 1:length(data$y))
    {
      sum_x = 0
      lambda = 1000*exp(-mu*(t+1))*psi;
      for (i in 1:y[t]){
        sum_x <- sum_x + i
      }
      ll <- ll - lambda + y[t]*log(lambda) - sum_x
    }
    ll
  }
)
```


### Obtain the data
Data used in the likelhood function can be in any format. 

```{r data import}

dataMos <- read.csv("./RWM_mosquito.csv")
# restructure the data for the log likelihood function
data <-   list(
  time = dataMos$time,
  y = dataMos$recaptured
)
```

### Settings
The settings for the parallel tempering calibrations are summarised here

#### Settings Options

* nrChains, number of independent chains to run
* M, number of dependent chains per chain run (i.e. the number of rungs in the temperature ladder)
* iterations, the number of steps to take in the Markov chain, (including the burn in)
* burn, the number of steps int he burn in (these are discarded)
* thin, thinning of the chain (i.e. =10 means only everything 10th sample is saved)
* consoleUpdates, frequency in which the console updates (i. =100 means every 100th step)
* P, number of parameters
* adap_Cov, whether to include adaptive covariance
* adap_Cov_freq, frquency the adaptive covariance matrix os updated
* adap_Cov_burn, number of steps to take before using the adaptive covariance matrix 
* adap_Temp, wheter to include adaptive temperature ladder
* adap_Temp_freq, frequency the adaptive temperature ladder is updated
* Debug, run with debug output 

```{r settings}

# settings used for the ptmc model (See README)
settings <-  list(
  nrChains = 3,
  M = 10,
  iterations = 10000, 
  burn = 1,
  thin = 1,
  consoleUpdates = 100,
  P = 2,        
  adap_Cov = TRUE,
  adap_Cov_freq = 1,
  adap_Cov_burn = 1000,
  adap_Temp = TRUE,
  adap_Temp_freq = 1,
  Debug = FALSE
)
```

### Run the model.

```{r run model,  message=FALSE, results = 'hide'}
post <- ptmc_func(model=model, data=data, settings=settings)
```

## Plot the data 
`ptmc_func` returns a list of length two. The first element is `post$mcmc` and is a mcmc or mcmc.list object (from the code package). I can plot these and calcualte convergence diagnostics using coda functions:

```{r plot outcomes}
# Plot trace plots and posterior distributions for $mcmcriors for a chain i
par(mar = rep(2, 4)) # just for plotting purposes
summary(post$mcmc)
plot(post$mcmc)

# Plot the Gelman-Rubin diagnostic for the parameters
gelman.plot(post$mcmc)
gelman.diag(post$mcmc)

```

The second element is `post$lpost` and is long table dataframe of the log posterior values. These values can be easily plotted using ggplot2:
```{r}
# Plot of the logposterior for the three chains
lpost_conv <- post$lpost %>% filter(sample_no>5000)
logpostplot <- ggplot(lpost_conv, aes(x = sample_no, y = lpost)) + 
  geom_line(aes(color = chain_no), size = 0.2, alpha=0.8) +
  theme_minimal()
logpostplot
```